---
title: "TheWholePat_2"
author: "Yuan"
date: "10/24/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
```

## Part 1: Data preparation
```{r load packages}
library(stringr)
library(plyr)
library(dplyr)
library(ggplot2)
library(GGally)
library(CCA)
library(pheatmap)
library(plotly)
library(reshape2)
library(PMA)
library(caret)
library(Hmisc)
library(parallel)
library(knitr)
# load the functions which are writed by Cedric Huchuan Xia
source('E:\\lab\\FEP\\1011sCCA\\cca_functions.R')
```

```{r Set the main path}
path = 'E:\\lab\\FEP\\FEP2'
```

```{r load data}
fn = 'FEP_MRI_list.csv'  
filename = paste(path,fn,sep = '\\')
clin = read.csv(filename)
colnames(clin)[1] = 'ID'
rownames(clin) = clin[,1]

fn = 'FEP_fa_Feb2018.csv' 
filename = paste(path,fn,sep = '\\')
fa = read.csv(filename)
colnames(fa)[1] = 'ID'
colnames(fa)[2:length(fa)] = paste("fa",colnames(fa[,c(2:length(fa))]))
rownames(fa) = fa[,1]

fn = 'FEP_MD_Feb2018.csv'  ### your data
filename = paste(path,fn,sep = '\\')
md = read.csv(filename)
colnames(md)[1] = 'ID'
colnames(md)[2:length(md)] = paste("md",colnames(md[,c(2:length(md))]))
rownames(md) = md[,1]

fn = 'FEP_fmri_Feb2018.csv'  ### your data
filename = paste(path,fn,sep = '\\')
fmri = read.csv(filename)
colnames(fmri)[1] = 'ID'
colnames(fmri)[2:length(fmri)] = paste("fmri",colnames(fmri[,c(2:length(fmri))]))
rownames(fmri) = fmri[,1]

fn = 'np.csv' 
filename = paste(path,fn,sep = '\\')
np = read.csv(filename)
np = subset(np,np$visit == 'year-1')

```

```{r concatenate multi-modal data}
mt = join_all(list(clin,fa,md,fmri), by = 'ID',type = 'inner')
mt[mt == ""] = NA
mt = mt[ , apply(mt, 2, function(x) !any(is.nan(x)))]
mt = subset(mt,mt$visit == 'year-1')  

mt = join_all(list(np,mt), by = 'study_ID',type = 'inner')
mt=mt[!is.na(mt$gds),]
#mt <- na.omit(mt)
mt=mt[-dim(mt)[1],] # remove FEP2068, repetitive subject
```

```{r seperate NC and Patients}
PAT = subset(mt,group == 1) #patients
NC = subset(mt,group == 2) #controls
```

```{r select fmri dataset in PAT and NC group}

fMRI_PAT=select(PAT,starts_with('fmri'))
rownames(fMRI_PAT)=PAT$study_ID

fMRI_NC=select(NC,starts_with('fmri'))
rownames(fMRI_NC)=NC$study_ID
```


```{r Select FS (6 Factor scores) for NC and patients}
FS_NC=select(NC,ends_with('_Std')) # 6 Factor Score
rownames(FS_NC)=NC$study_ID
FS_PAT=select(PAT,ends_with('_Std')) # 6 Factor Score
rownames(FS_PAT)=PAT$study_ID
```



## Part 2: Do sCCA
```{r creat sample }
sampleid = 1:nrow(FS_PAT)

brain_sample <- mclapply(sampleid, function(id) fMRI_PAT[id,])
behavior_sample <- mclapply(sampleid, function(id) FS_PAT[id,])
```

### 2.1 select best parameters for regulariation
```{r }
#which(is.na(FS_PAT))#There were 2 NA in FS_PAT
#replace the NA by mean value
FS_PAT[,1] = impute(FS_PAT[,1],mean)
FS_PAT[,5] = impute(FS_PAT[,5],mean)
x_pen <- seq(0.1,1,length.out=10)
y_pen <- seq(0.1,1,length.out=10)
gs.mat <- matrix(0, nrow = 10, ncol = 10)

for (pen_xseq in x_pen) {
  pen_xseq_int = round(pen_xseq*10)
  for (pen_yseq in y_pen) {
    pen_yseq_int = round(pen_yseq*10)
    gs=ccaDW(fMRI_PAT, FS_PAT,pen_xseq,pen_yseq,6)
    gs.mat[pen_xseq_int,pen_yseq_int] = mean(abs(gs$cor))
  }
}
```
```{r find the best parameters }
rownames(gs.mat) <- x_pen
colnames(gs.mat) <- y_pen
pheatmap(gs.mat,cluster_cols = FALSE,cluster_rows = FALSE,treeheight_col = FALSE,treeheight_row = FALSE)
which.max.matrix(gs.mat)#select the best parameters of max coralations
```


### 2.2 select components based on covariance matrix
```{r }
modenum <- dim(FS_PAT)[2] #number of all possible canonical variates
scca.org <- ccaDW(fMRI_PAT, FS_PAT,0.8,0.9,modenum) #0.8 and 0.9 here are the best parameteres selected above in the grid search
brain_std = apply(fMRI_PAT, 2, scale)
med_std <- apply(FS_PAT,2,scale)
covmat <- t(scca.org$u) %*% t(brain_std) %*% med_std %*% scca.org$v #calculate covariance matrix
varE <- diag(covmat)^2 / sum(diag(covmat)^2) #calcualte covariance explained by each component
varE.df <- data.frame(modenum = as.factor(1:modenum), var = varE) #prepare dataframe for plotting
candnum = 6 #number selected based on the scree plot

p.var<-ggplot(varE.df,aes(modenum,var)) +
  geom_point(stat = 'identity',aes(color = var>0.1, size = 10)) +
  geom_hline(yintercept = 1/10,linetype="dashed") +
  scale_x_discrete(name ="Canonical-Var", limits=c(1:modenum),breaks =  c(1,seq(1,modenum,1))) +
  scale_y_continuous(expand = c(0, 0),limits=c(0,1),labels = scales::percent,name = "Variance Explained", breaks=seq(0,1,length=4)) +
  theme_classic(base_size = 20) +
  theme(legend.position = 'none') 

p.var
```


```{r run sCCA on each sample to estimate average correlation}
if(FALSE){
scca.cand <- ccaDW(fMRI_PAT, FS_PAT,0.8,0.9,candnum)
scca.cca<-mclapply(sampleid,function(i) ccaDW(brain_sample[[i]],behavior_sample[[i]],0.8,0.9,6)) #loop through split
scca.cca.ro <- sapply(scca.cca,function(x) reorderCCA(x,scca.cand,6)) #reorder the component so they match across splits
scca.cca.cor <- rowMeans(simplify2array(scca.cca.ro['cors',]),na.rm =T) #calculate average of cca correlations
scca.cca.cor.se <- rowSds(simplify2array(scca.cca.ro['cors',]),na.rm =T)/sqrt(dim(scca.cca.ro)[2]) #calculate standard error of correlations  
}
```


### 2.3 permutation test
```{r permutation test}
scca.cand <- ccaDW(fMRI_PAT, FS_PAT,0.8,0.9,candnum)
num.perm <- 1000 #number of permutaitons to run
behavior.perm <- rlply(num.perm,FS_PAT[sample(nrow(FS_PAT)),]) #permute the clinical matrix by row
scca.perm.cca<-sapply(behavior.perm, function(y_perm){ out<-ccaDWpermorder(fMRI_PAT,y_perm,0.8,0.9,candnum,scca.cand)} ) #run scca again with permuted clinical but with original connectivity
perm.cor <- simplify2array(scca.perm.cca['cors',]) #extrac the correlations
perm.pval = array(0,dim = 6)
```

```{r calculate p-value}
for (f in 1:6) {
  perm.cor.cv =  perm.cor[f,]
  perm.pval[f] = length(which(perm.cor.cv>=scca.cand$cors[f]))/length(which(is.na(perm.cor.cv) == FALSE))
}
perm.pval
scca.org$cors
```

```{r calculate the scores}
brain_scores = brain_std %*% scca.org$u
behavior_scores = med_std %*% scca.org$v
v = scca.org$v#cognitive factor loadings 
rownames(v) = colnames(FS_PAT)
kable(v)
```
  
We can see that the first canonical variate is Memory-related  
the second canonical variate is Ideational-related  
the 6th canonical variate is IdeationalExecutive-related(which is significant through permutation test)


### 2.4 scatter plot about correlation
```{r }
df1=cbind(as.data.frame(behavior_scores[,1]),as.data.frame(brain_scores[,1]))
colnames(df1)=c('cognition','connectivity')
ggplot(df1, aes(x=cognition, y=connectivity))+ 
  geom_point( size = 2,alpha=0.7)+theme_bw(base_size=15)+geom_smooth(method = "lm",se = FALSE )+
  labs(x='cognitive score',y='connectivity score',title="Correlation: Memory-related")
```

```{r }
df1=cbind(as.data.frame(behavior_scores[,2]),as.data.frame(brain_scores[,2]))
colnames(df1)=c('cognition','connectivity')
ggplot(df1, aes(x=cognition, y=connectivity))+ 
  geom_point( size = 2,alpha=0.7)+theme_bw(base_size=15)+geom_smooth(method = "lm",se = FALSE )+
  labs(x='cognitive score',y='connectivity score',title="Correlation: Ideational-related")
```
```{r }
df1=cbind(as.data.frame(behavior_scores[,6]),as.data.frame(brain_scores[,6]))
colnames(df1)=c('cognition','connectivity')
ggplot(df1, aes(x=cognition, y=connectivity))+ 
  geom_point( size = 2,alpha=0.7)+theme_bw(base_size=15)+geom_smooth(method = "lm",se = FALSE )+
  labs(x='cognitive score',y='connectivity score',title="Correlation: IdeationalExecutive-related")
```
  
We can see that:
the second canonical variate is not so much related between cognitive score and connectivity score  
so,we keep the 1st(which has most Explained Variance) and 6th canonical variates(which is significant) to do clustering

## Part 3: Clustering

### 3.1 Hierachical clustering
```{r}
CV_cog = cbind(as.data.frame(behavior_scores[,1]),as.data.frame(behavior_scores[,6]))
colnames(CV_cog) = c('Memory-related','IdeationalExecutive-related')

d = dist(scale(CV_cog),method = 'euclidean')
hc = hclust(d,method = 'ward.D')
clust = cutree(hc,k = 4)
table(clust)
```

### 3.2 Heat map
```{r}
clust=as.factor(clust)
PAT_whole=cbind(clust,PAT)
mat=t(as.matrix(CV_cog))
cv=data.frame('Diag'=as.factor(PAT_whole$diag2),'Cluster'=as.factor(PAT_whole$clust))
rownames(cv)=PAT_whole$study_ID
colnames(cv)=c('Diag','Cluster')
phem=pheatmap(mat,scale = 'row',annotation_col = t(cv), clustering_method="ward.D",cutree_cols=4,
              clustering_distance_cols = "euclidean",show_colnames = TRUE)
```

#### 3.3 Scatter plot

make a scatterplot for four clusters of subjects along IdeationalExecutive-related and Memory-related fmri scores 
```{r}
CV_cog2=cbind(CV_cog,clust)
ggplot(CV_cog2, aes(x=`Memory-related`, y=`IdeationalExecutive-related`, col=clust)) +
  geom_point(aes(shape=clust), size=2)+
  coord_cartesian(xlim = c(-3,3), ylim = c(-3,3))  + 
  geom_segment(x = 0,xend=0, y = -3,yend=3,size=1,col='black')+
  geom_segment(x = -3,xend=3, y = 0,yend=0,size=1,col='black')+theme_minimal()+
  labs(title='Memory- and IdeationalExecutive-related connectivity for 4 clusters')
```

#### 2.2.4 Pattern of abnormal connectivity 
Calulate fisher Z-score of fmri connectivity for  NC group
```{r}
n=nrow(fMRI_NC) #number of subjects
fMRI_labels=read.table(str_c(path,'\\fMRI_label.txt'))
q=ncol(fMRI_labels) # number of parcels for fMRI
fMRI_matrix=matrix(seq(1,q^2),ncol = q,nrow=q) # initialize 2D distance matrix for each subject
idx_lower=fMRI_matrix[lower.tri(fMRI_matrix)]
rownames(fMRI_matrix)=as.matrix(fMRI_labels)
colnames(fMRI_matrix)=as.matrix(fMRI_labels)
fMRI_array=array(0,dim=c(q,q,n)) # initialize 3D distance matrix for all subjects, the 3rd dimension corresponds to subject index
for (j in 1:n){
count=1
fMRI_matrix=matrix(0,ncol = q,nrow=q)
for (i in idx_lower) {
  fMRI_matrix[i]=fMRI_NC[j,count]
  count=count+1
}
fMRI_matrix=t(fMRI_matrix)+fMRI_matrix
fMRI_array[,,j]=fMRI_matrix
}
```

calculate the mean value of fMRI matrix in NC group(95 subjects)
```{r}
fMRI_matrix_mean = matrix(seq(1,q^2),ncol = q,nrow=q) # initialize 2D distance matrix 
  for (j in 1:78){
    for (k in 1:78){
      fMRI_matrix_mean[j,k] = mean(fMRI_array[j,k,])
    }
  }
```


Here We generate average matrix showing normal connecivtity. Each element is Z score of U statistic of U-test.

```{r}
fMRI_Utest_vec=vector(length = ncol(fMRI_NC)) # initialize 1D vector
for (i in 1:ncol(fMRI_NC))
{
  Utest=wilcox.test(fMRI_PAT[,i], fMRI_NC[,i])
  Umean=nrow(fMRI_NC)*nrow(fMRI_PAT)/2
  Usd=sqrt(nrow(fMRI_NC)*nrow(fMRI_PAT)*(nrow(fMRI_NC)+nrow(fMRI_PAT)+1)/12)
  Z=(Utest$statistic-Umean)/Usd
  fMRI_Utest_vec[i]=Z
}
fMRI_Utest_mat=matrix(seq(1,q^2),ncol = q,nrow=q) # initialize 2D distance matrix
idx_lower=fMRI_Utest_mat[lower.tri(fMRI_Utest_mat)]
count=1
fMRI_Utest_mat=matrix(0,ncol = q,nrow=q)
for (i in idx_lower) {
  fMRI_Utest_mat[i]=fMRI_Utest_vec[count]
  count=count+1
}
fMRI_Utest_mat=t(fMRI_Utest_mat)+fMRI_Utest_mat
```

Plot the matrix above
```{r}
fMRI_labels2=read.table(str_c(path,'\\fMRI_label2_2.txt'),stringsAsFactors=FALSE)#this file is the result of clustering percels
rownames(fMRI_Utest_mat)=as.matrix(fMRI_labels)   
colnames(fMRI_Utest_mat)=as.matrix(fMRI_labels)  
cv_fmri=data.frame('group'=fMRI_labels2$V1)
rownames(cv_fmri)=as.matrix(fMRI_labels) 
### order the fMRI_Utest_mat by groups(BrainNet)
index <- order(cv_fmri[,1])
fMRI_Utest_mat_reorder = fMRI_Utest_mat[index,index]
fMRI_labels2_reorder = cv_fmri[index,]

col4 <- colorRampPalette(c("#00ffff","#0066ff","black","black","black","black","black","orange","red","yellow"))
#fMRI_labels2$V1=as.factor(fMRI_labels2$V1)
library(RColorBrewer)
mat_colors_fmri <- list(group = brewer.pal(length(levels(fMRI_labels2_reorder)), "Paired"))
# color palettes: http://www.datavis.ca/sasmac/brewerpal.html
# display.brewer.all() # show all color palettes
names(mat_colors_fmri$group) <- unique(fMRI_labels2_reorder) 
gaps_vec=vector(length =length(unique(fMRI_labels2_reorder) )-1 )
for (i in 1:(length(unique(fMRI_labels2_reorder) )-1))
{
  gaps_vec[i]=which(fMRI_labels2_reorder==as.character(unique(fMRI_labels2_reorder)[i+1]))[1]
}


breaksList =seq(-4, 6, by = 1)


pheatmap(
  mat               = fMRI_Utest_mat_reorder,cluster_row = F, cluster_col = F,
  color             = col4(10),
  breaks=breaksList,
  border_color      = NA,
  show_colnames     = FALSE,
  show_rownames     = FALSE,
  annotation_col    = cv_fmri,
  annotation_colors = mat_colors_fmri,
  drop_levels       = TRUE,
  fontsize          = 14,
  gaps_col          = gaps_vec-1,
  gaps_row          = gaps_vec-1,
  main              = 'Pattern of abnormal connectivity shared by BrainNet'
)
```

### 3.5 Pattern of abnormal connectivity in 4 clusters(PAT group)
Here We generate a matrix showing abnormal connecivtity for each biotype. Each element is Z score of U statistic of U-test.
plot Biotype 1
```{r}
PAT1=filter(PAT_whole,clust==1)
fMRI_PAT1=select(PAT1,starts_with('fmri'))
rownames(fMRI_PAT1)=PAT1$study_ID

col4 <- colorRampPalette(c("#00ffff","#0066ff","black","black","black","black","black","orange","red","yellow"))

fMRI_Utest_vec1=vector(length = ncol(fMRI_NC)) # initialize 1D vector
for (i in 1:ncol(fMRI_NC))
{
  Utest=wilcox.test(fMRI_PAT1[,i], fMRI_NC[,i])
  Umean=nrow(fMRI_NC)*nrow(fMRI_PAT1)/2
  Usd=sqrt(nrow(fMRI_NC)*nrow(fMRI_PAT1)*(nrow(fMRI_NC)+nrow(fMRI_PAT1)+1)/12)
  Z=(Utest$statistic-Umean)/Usd
  fMRI_Utest_vec1[i]=Z
}

count=1
fMRI_Utest_mat1=matrix(0,ncol = q,nrow = q)
for (i in idx_lower) {
  fMRI_Utest_mat1[i]=fMRI_Utest_vec1[count]
  count=count+1
}
fMRI_Utest_mat1=t(fMRI_Utest_mat1)+fMRI_Utest_mat1
breaksList =seq(-4, 6, by = 1)
rownames(fMRI_Utest_mat1)=as.matrix(fMRI_labels)
colnames(fMRI_Utest_mat1)=as.matrix(fMRI_labels)
fMRI_Utest_mat1_reorder = fMRI_Utest_mat1[index,index]
pheatmap(
  mat               = fMRI_Utest_mat1_reorder,cluster_row = F, cluster_col = F,
  color             = col4(10),
  breaks=breaksList,
  border_color      = NA,
  show_colnames     = FALSE,
  show_rownames     = FALSE,
  annotation_col    = cv_fmri,
  annotation_colors = mat_colors_fmri,
  drop_levels       = TRUE,
  fontsize          = 14,
  gaps_col          = gaps_vec-1,
  gaps_row          = gaps_vec-1,
  main              = 'Pattern of abnormal connectivity for biotype 1'
)
```


plot Biotype 2
```{r}
PAT2=filter(PAT_whole,clust==2)
fMRI_PAT2=select(PAT2,starts_with('fmri'))
rownames(fMRI_PAT2)=PAT2$study_ID

fMRI_Utest_vec2=vector(length = ncol(fMRI_NC)) # initialize 1D vector
for (i in 1:ncol(fMRI_NC))
{
  Utest=wilcox.test(fMRI_PAT2[,i], fMRI_NC[,i])
  Umean=nrow(fMRI_NC)*nrow(fMRI_PAT2)/2
  Usd=sqrt(nrow(fMRI_NC)*nrow(fMRI_PAT2)*(nrow(fMRI_NC)+nrow(fMRI_PAT2)+1)/12)
  Z=(Utest$statistic-Umean)/Usd
  fMRI_Utest_vec2[i]=Z
}

count=1
fMRI_Utest_mat2=matrix(0,ncol = q,nrow=q)
for (i in idx_lower) {
  fMRI_Utest_mat2[i]=fMRI_Utest_vec2[count]
  count=count+1
}
fMRI_Utest_mat2=t(fMRI_Utest_mat2)+fMRI_Utest_mat2

rownames(fMRI_Utest_mat2)=as.matrix(fMRI_labels)
colnames(fMRI_Utest_mat2)=as.matrix(fMRI_labels)
fMRI_Utest_mat2_reorder = fMRI_Utest_mat2[index,index]
pheatmap(
  mat               = fMRI_Utest_mat2_reorder,cluster_row = F, cluster_col = F,
  color             = col4(10),
  breaks=breaksList,
  border_color      = NA,
  show_colnames     = FALSE,
  show_rownames     = FALSE,
  annotation_col    = cv_fmri,
  annotation_colors = mat_colors_fmri,
  drop_levels       = TRUE,
  fontsize          = 14,
  gaps_col          = gaps_vec-1,
  gaps_row          = gaps_vec-1,
  main              = 'Pattern of abnormal connectivity for biotype 2'
)
```



plot Biotype 3
```{r}
PAT3=filter(PAT_whole,clust==3)
fMRI_PAT3=select(PAT3,starts_with('fmri'))
rownames(fMRI_PAT3)=PAT3$study_ID

fMRI_Utest_vec3=vector(length = ncol(fMRI_NC)) # initialize 1D vector
for (i in 1:ncol(fMRI_NC))
{
  Utest=wilcox.test(fMRI_PAT3[,i], fMRI_NC[,i])
  Umean=nrow(fMRI_NC)*nrow(fMRI_PAT3)/2
  Usd=sqrt(nrow(fMRI_NC)*nrow(fMRI_PAT3)*(nrow(fMRI_NC)+nrow(fMRI_PAT3)+1)/12)
  Z=(Utest$statistic-Umean)/Usd
  fMRI_Utest_vec3[i]=Z
}

count=1
fMRI_Utest_mat3=matrix(0,ncol = q,nrow=q)
for (i in idx_lower) {
  fMRI_Utest_mat3[i]=fMRI_Utest_vec3[count]
  count=count+1
}
fMRI_Utest_mat3=t(fMRI_Utest_mat3)+fMRI_Utest_mat3

rownames(fMRI_Utest_mat3)=as.matrix(fMRI_labels)
colnames(fMRI_Utest_mat3)=as.matrix(fMRI_labels)
fMRI_Utest_mat3_reorder = fMRI_Utest_mat3[index,index]
pheatmap(
  mat               = fMRI_Utest_mat3_reorder,cluster_row = F, cluster_col = F,
  color             = col4(10),
  breaks=breaksList,
  border_color      = NA,
  show_colnames     = FALSE,
  show_rownames     = FALSE,
  annotation_col    = cv_fmri,
  annotation_colors = mat_colors_fmri,
  drop_levels       = TRUE,
  fontsize          = 14,
  gaps_col          = gaps_vec-1,
  gaps_row          = gaps_vec-1,
  main              = 'Pattern of abnormal connectivity for biotype 3'
)
```


plot Biotype 4
```{r}
PAT4=filter(PAT_whole,clust==4)
fMRI_PAT4=select(PAT4,starts_with('fmri'))
rownames(fMRI_PAT4)=PAT4$study_ID

fMRI_Utest_vec4=vector(length = ncol(fMRI_NC)) # initialize 1D vector
for (i in 1:ncol(fMRI_NC))
{
  Utest=wilcox.test(fMRI_PAT4[,i], fMRI_NC[,i])
  Umean=nrow(fMRI_NC)*nrow(fMRI_PAT4)/2
  Usd=sqrt(nrow(fMRI_NC)*nrow(fMRI_PAT4)*(nrow(fMRI_NC)+nrow(fMRI_PAT4)+1)/12)
  Z=(Utest$statistic-Umean)/Usd
  fMRI_Utest_vec4[i]=Z
}

count=1
fMRI_Utest_mat4=matrix(0,ncol = q,nrow=q)
for (i in idx_lower) {
  fMRI_Utest_mat4[i]=fMRI_Utest_vec4[count]
  count=count+1
}
fMRI_Utest_mat4=t(fMRI_Utest_mat4)+fMRI_Utest_mat4

rownames(fMRI_Utest_mat4)=as.matrix(fMRI_labels)
colnames(fMRI_Utest_mat4)=as.matrix(fMRI_labels)
fMRI_Utest_mat4_reorder = fMRI_Utest_mat4[index,index]

pheatmap(
  mat               = fMRI_Utest_mat4_reorder,cluster_row = F, cluster_col = F,
  color             = col4(10),
  breaks=breaksList,
  border_color      = NA,
  show_colnames     = FALSE,
  show_rownames     = FALSE,
  annotation_col    = cv_fmri,
  annotation_colors = mat_colors_fmri,
  drop_levels       = TRUE,
  fontsize          = 14,
  gaps_col          = gaps_vec-1,
  gaps_row          = gaps_vec-1,
  main              = 'Pattern of abnormal connectivity for biotype 4'
)
```